---
title: "DataFrame"
author: "Taiwo"
date: "3/1/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Frames

A data frame is a table or a two-dimensional array
like structure in which each column contains values of one variable and each row contains one set of values from each column.

## Charateristics of a Data Frame

The following are the characteristics of a Data Frame:

*  The column names should be non-empty
*  The row names should be unique 
*  The data stored in a data frame can be numeric, factor or character type.
*  Each column should contain same number of data items.

## Creating a Data Frames

To create a data frame, we shall use the data frame function called data.frame(). 

For example, this code creates a data frame of five employees information with their respective names,
age, salaries, profession.

```{r}
employees_info <- data.frame(
  emp_names=c("Frank", "Micheal", "Peter", "Melissa", "Esther"),
  emp_age = c(43, 32, 40, 41, 45), 
  salaries = c(100000, 200000, 50000, 500000, 70000),
  emp_profession = c("Data Scientist", "Senior It Admin", "associate accountant","Excutive director", "junior data scientist"), 
  stringsAsFactors = FALSE
)
print(employees_info)
```


Also, to find class of an object, use class() function. For example, the class of employees_info object is `r class(employees_info)`.

Additionally, the structure of the the data frame 
could be found using the str() function. For example, we could find the structure of the dataframe created from the previous example like this:
```{r}
str(employees_info)
```


## Summary of Variables in Data Frame

The summary() function was applied here to encapsulate each variable in the Data Frame. As an example, apply summary() function to the employee_info data frame created earlier. While emp_age and salaries variables are numerical variables, emp_names and emp_profession are character variables.    
```{r}
 summary(employees_info)
```


## Manipulating column names and row names

To find the row names of any data frame, apply the 
rownames() function to data frame. For example, Using the employees_info earlier created, rownames(employees_info) returns the names for each row in the data frame, employees_info.
```{r}
rownames(employees_info)
```

Also, we can use rownames() function to change add row names. For example, we can add row names to employee_info using the following method
```{r}
rownames(employees_info) <- c("row1", "row2", "row3", "row4", "row5")
employees_info
```

Furthermore, colnames() function can be applied to any data frame to return name of each variable. Accordingly, the following code returns variable names for each column in the employee_info data frame. Also, we could use dimnames() function as well.
```{r}
colnames(employees_info)
dimnames(employees_info)
```

Just like the way we add row names to each row, we could change column's names to another name using colnames() function like the following example:
```{r}
colnames(employees_info) <- c("employee_names", "employee_age", "employee_salaries")
employees_info
```


\newpage

## Indexing Data Frames
Elements in data frames are selected using data frame's indexing. For instance, we could select the 
first and second rows for the first and the second column.
```{r}
employees_info[1:2, 1:2]
```

It is possible to extract row's elements with a single column. In the following example, the first, second and third elements of the first column was extracted from the employees_info data frame.
```{r}
employees_info[[1]][1:3]
```


# cbind() and rbind() with data frame.

The meaning of cbind() is column bind. cbind() function is used to combine vectors, matrices, and data frames by columns. As case in point, the code that follows applied cbind() function to combine data_info with v3 vector by column to form a data frame.
```{r}
#Create a dataframe
data_info <- data.frame(
col1 = 1:3,
col2 = c(2, 3, 4),
col3 = c(120, 130, 150)) 

#create a vector of decimal number 
v3 <- c(12.23, 22.20,  30.30) 

#Combine a dataframe with the vector using cbind() function
cbind(data_info, v3)
```

In the same vein, the name rbind means row-bind. The rbind() function can be used to combine vectors, matrices and/or data frames by row. However, caution must be observed when applying rbind() function for combination because this can cause changes in the columns classes if not handle properly. For instance, if we were to combine the data_info with v3, this will change all column class to a float class. In order to avoid this, we will not be using a vector but rather a data frame in this form:
```{r}
df <- data.frame(col1 = 4,
                 col2 = 5,
                 col3 = 200)
rbind(data_info, df)
```


# Adding Attibutes to Data Frames

```{r}

```

