---
title: "List"
author: "Taiwo Famuyiwa"
output:
  pdf_document: default
  html_document:
    df_print: paged
---


```{r, include=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE, comment = NA)
```


\begin{center}
How to create a List in R
\end{center}

A list is a vector where each element can be of a 
different type. Lists are created with the list function. List contents are created, with each argument separated by a comma. List elements can be any variable type - vectors, matrices etc. 

In the following example, a list called list_data is created. There are three elements in the list, each one with different data type, these are: list, vector and a matrix.
```{r collapse=TRUE}
(list_data = list(c(1, 2, 3, 4), c("taiwo", "bunmi", "wunmi", "Funmi"), matrix(c(1, 2, 3, 4), nrow = 2)))
```


## Naming Each element of a List
Like vectors, it is possible to give name to each element of a list during the construction, or afterward using the names function.In the example that follows, name was given to each element in the list, using the name function. 
```{r collapse=TRUE}
list1_element <- list(c("Monday", "Tuesday", "Wednesday"), c("January", "February", "March"), c(2001, 2002, 2003))
names(list1_element) <- c("Days", "Month", "year")
list1_element
```


\newpage
## Nested List
Lists can be nested. The following example demonstrates this. Here, two lists were nested within a list
using the list function.
```{r collapse=TRUE}
(account_list <- list(
  sales_product = list(product = c("butter", "bread", "beans"), sales = c("$20", "$30", "$40")),  cost_produce = 
    list(product = c("butter", "bread", "beans"),
         cost = c("$10", "$20", "$300"))
))
```

## Atomic and Recursive Variables

A  list is called recursive variable because it contains other list within itself. While vectors, matrices and arrays are atomic, a list stands out from the rest because it is recursive variable.
In the following examples, it shows that a list is recursive not atomic. 

```{r collapse=TRUE}
is.atomic(list())
is.recursive(list())
is.atomic(c(1, 2, 3, 4))
is.recursive(c(1, 2, 3, 4))
```

## List Dimensions and Aritmetic

A list have length. A list length is the number of 
top level elements that it contains.

```{r collapse=TRUE}
list_data = list(c(1, 2, 3, 4), c("taiwo", "bunmi", "wunmi", "Funmi"), 
                  matrix(c(1, 2, 3, 4), nrow = 2))

length(list_data)
```


\newpage
## Accessing List Elements
There are three ways to access the element of a list and they are as follows:

*  Access the element of a list by its element's name e.g list_1$c
*  Access the element of a list by its element's name inside a bracket. e.g list_1[[c]]
*  Access the element index i of the element within a list. e.g list_2[[i]]

Furthermore, we could access the element of a list using a single bracket.
For instance, list_2["c"] and list_2[i]. Both double bracket and single bracket 
access elements in vector-index fashion. The only difference is while single 
bracket index return a list - sublist of the original, double bracket return 
element in terms of a vector. The following code demonstrates the importance 
of this illustration. The first and the third examples return a vector while the
second return a vector. Also, the fourth example return a matrix. 
```{r collapse=TRUE}
list_data = list(c(1, 2, 3, 4), c("taiwo", "bunmi", "wunmi", "Funmi"), 
                  matrix(c(1, 2, 3, 4), nrow = 2))
names(list_data) <- c("vector", "number", "matrix")
list_data[["vector"]] #Return a vector 
list_data["vector"] #Return a list
list_data[[2]]  #This will return the second element of a list by indexing using a number 
list_data$matrix  #Index element of a list by name
```


## Adding and Deleting List Elements 

It is possible to add an element to a list after it has been created.The 
following example demonstrates that. For example the, the second indexing sees 26 
added to the list using the index method. Also, multiple elements can also be
added to the list using multiple indexing. In the last example, the last element
of a list is removed using the NULL statement.
```{r collapse=TRUE}
z <- list(a = "abc", b = 12)
#To add element to a list, do the following
z$c <- "He was there yesterday"
#Adding element to a list can also be done via a vector index 
z[[4]] <- 26
#It is also possible to add multiple elements at the same time to a list
z[5:7] <- c(12, 33, 44)
#The following code delete an element from a list.
z$c <- NULL
```


\newpage
## Converting Between Vectors and Lists

One of the advantages of R language is that it is possible to convert a vector
to a list. This can be achieved by applying the as.list() function. This creates
a list with each element of the vector mapping to a list element containing one 
value. 
```{r collapse=TRUE}
vector_one <- c(2, 3, 4, 5, 6)
as.list(vector_one)
```


It is possible to revert the operation by converting the list to a vector
using as.numeric() function respectively. In the example that follows, the list 
that was created in the previous example is converted to a numeric vector using
the as.numeric() function.
```{r collapse=TRUE}
as.numeric(vector_one)
```

Additionally, the unlist() function can be use to convert a list to  a vector.
For instance, in the following code, unlist() function transform a list to a 
vector.
```{r collapse=TRUE}
list_one <- list(x = 12, y=23, z=24)
unlist(list_one)
```

## Combining Lists

Just like the way the c function was used to concatenate vectors, it can be use
to concatenate two or more lists. In the example that follows, the c function 
concatenate two lists.
```{r collapse=TRUE}
c(list(x = 3, y = 4, z = 5), list(7))
```

In addition,the c function can be use to concatenate a list and a vector 
together. This will convert the vector to a list.
```{r collapse=TRUE}
c(list(x = 3, y = 4), 3)
```


## Applying Functions to Lists
We could apply lapply() and sapply() functions to lists. It is possible to call
lapply() function on each component of a list and returns a list.
The following example elucidate this method. Here, R applied median() to 1:6 and
20:40, returning a list of median number for each component of a list.
```{r collapse=TRUE}
lapply(list(1:6, 20:40), median)
```

sapply() is different from lapply() because it returns a vector instead of 
a list. For instance, apply sapply() function to the previous example returns 
the vector of numbers instead of a list.
```{r collapse=TRUE}
sapply(list(1:6, 20:40), median)
```

## NULL 

Another important special value that must be mentioned is empty variable.
This type of special number is prominent while working with list, data frames or 
during function arguments (we shall be talking about in subsequent topic).

For instance, while creating a list, it is possible to create an element without 
contents. A NULL special number can be used to do that. For instance, the 
following list income per cities in some states in the year 2000. 
Here we specify that some cities have an income while others do not.
```{r collapse=TRUE}
(profit_cities <- list(
  Texas = "one million dollars", 
  dallas = "two million dollars", 
  little_rock = NULL, 
  new_york = NULL
  ))
```

\newpage
It is important to understand the difference between NULL and special missing
value NA. The biggest difference is that NA is a scalar value, whereas NULL 
takes up no space at all - that is length zero. For instance, using length() 
function, the length of NA is 1 while the length of NULL is 0. Also, to test
for NULL, apply is.null() function. This will return a logical statement.
```{r collapse=TRUE}
#The length NULL is 0
length(NULL)

#The length of NA is 1 
length(NA)

#Apply is.null()
is.null(NULL)
```

Furthermore, NULL statement can be used to remove elements of a list. Setting an
element to NULL will remove it. For instance, using the previous examples
we could set texas element to NULL. This will remove the element from the list.
```{r collapse=TRUE}
profit_cities$Texas <- NULL
profit_cities
```

